Type Evolution in Interactive Programming Systems

### 2024-05-31 paper.tex
more tweaks

The goals for this challenge are twofold: from the end-user perspective and from the language engineering perspective. A particular challenge from the end-user perspective is to ``do minimal harm'': it is essential for fluid programming experience that the automatically triggered migrations are in a sense as near as possible to the previous application state, to not surprise or confuse the user/developer.
One approach  considered in earlier work~\cite{RuntimeConstraint} employs the constraint solver Z3 to find a ``nearest'' run-time instance compatible with a source change. Nevertheless, the patching of the run-time state should be quick enough so as to not disrupt the programmer experience.

From the language engineering perspective the goal is to employ techniques, formalisms, and tools, to make the construction of such languages easier. The above example state machine DSL is derived from earlier work~\cite{vanRozen19}, where the authors manually implemented the run-time patch operation and concluded that even for such a simple language (simpler even than the example above) it is a complex and error-pone endeavor. Furthermore, the field of software language engineering studies and develops generic and reusable techniques to improve the development of DSLs and programming languages, for instance, in the context of language workbenches~\cite{ERDWEG201524}. The development of live programming languages, however, is currently still out of reach for all existing language workbenches. The aforementioned approach using a constraint solver is an example of such a \textit{language parametric} technique, in that it operates on the (extended) abstract syntax metamodel of a language, and does not assume anything further about the language itself. Another approach is the \textsc{Cascade} metamodeling formalism which has builtin support for run-time patching~\cite{Cascade}. However, further research is needed to design better principled language engineering approaches that solve the problem in a way that is both declarative and fast.

@@

While the above example is arguably simple, the problem becomes much more challenging when the programs themselves define data types, classes, records, structures etc. Since possibly many instances (values, objects) of such data types may exist at run-time, these all have to be migrated in such a way that programmer experience is minimally disrupted, and that the invariants of said data types is maintained.
Another extension, tying in with the data-oriented examples above, involves refactorings of data types in a program. Typically a refactoring should be behavior preserving, but can it also preserve run-time data? The minimal example is the consistent variable rename rename in Table~\ref{tbl:smchanges}, which should not have any effect on run-time state.  A more complex refactoring is described in Section~\ref{sec:elm}.

>>>>

The goals for this challenge are twofold: from the end-user perspective and from the language engineering perspective. A particular challenge from the end-user perspective is to ``do minimal harm'': it is essential for fluid programming experience that the automatically triggered migrations are in a sense as near as possible to the previous application state, to not surprise or confuse the user/developer.
% One approach  considered in earlier work~\cite{RuntimeConstraint} employs the constraint solver Z3 to find a ``nearest'' run-time instance compatible with a source change. Nevertheless, the patching of the run-time state should be quick enough so as to not disrupt the programmer experience.

From the language engineering perspective the goal is to employ techniques, formalisms, and tools, to make the construction of such languages easier. The above example state machine DSL is derived from earlier work~\cite{vanRozen19}, where the authors manually implemented the run-time patch operation and concluded that even for such a simple language (simpler even than the example above) it is a complex and error-pone endeavor. Furthermore, the field of software language engineering studies and develops generic and reusable techniques to improve the development of DSLs and programming languages, for instance, in the context of language workbenches~\cite{ERDWEG201524}. The development of live programming languages, however, is currently still out of reach for all existing language workbenches.

The above case study is concerned with changes to the DSL program (e.g., the Doors statemachine). However, there is another level of schema change to be considered: what if the metamodel (cf. Figure~\ref{FIG:statemachineUML}) itself is edited? In this case both the programs (e.g., Doors) possibly need to be migrated (in case of a change that is not backwards compatible), \textit{and} the running instance, including the interpreter, since its structure is, too, governed by the abstract syntax of the language.

%The aforementioned approach using a constraint solver is an example of such a \textit{language parametric} technique, in that it operates on the (extended) abstract syntax metamodel of a language, and does not assume anything further about the language itself. Another approach is the \textsc{Cascade} metamodeling formalism which has builtin support for run-time patching~\cite{Cascade}. However, further research is needed to design better principled language engineering approaches that solve the problem in a way that is both declarative and fast.

@@

While the above example is arguably simple, the problem becomes much more challenging when the programs themselves define data types, classes, records, structures etc. Since possibly many instances (values, objects) of such data types may exist at run-time, these all have to be migrated in such a way that programmer experience is minimally disrupted, and that the invariants of said data types is maintained.
Another extension, tying in with the data-oriented examples above, involves refactorings of data types in a program. Typically, a refactoring should be behavior preserving, but can it also preserve run-time data? The minimal example is the consistent variable rename in Table~\ref{tbl:smchanges}, which should not have any effect on run-time state.  A more complex refactoring is described in Section~\ref{sec:elm}.
